Bridging the Gap Between Visual and Analytical Machine Learning Testing

### 2023-10-25 report/icst24.tex
highlight asserts
add collection date
feat(report): 2023-10-25 13:54 updates

  Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate of the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part. Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems. And help ML practitioners formulate better analytical assertions.

@@

% TODO make sure that we use the terminology of "Testing Tasks" everywhere for RQ2
Visualisations are used throughout the ML development lifecycle to test various properties of the ML system. In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties. During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models. Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.

@@

We manually analysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation. Listing~\ref{lst:prompt} shows the prompt used for such cases. In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair. Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.

@@

@@

\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq2-p3}

@@

The results for RQ1 indicate that the combination of machine learning visualisations with analytical assertions is an emerging testing practice within the machine learning community. We found only 54,000 notebooks that contain the keyword ``assert'' and an image, while there are more than a million Jupyter notebooks available on GitHub. This suggests that the practice of assertion is still relatively uncommon.
We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties. We argue that we need more guidelines on ML coding practices and that education and research of software engineering practices ought to be steered towards the specific challenges entailed from contributing to ML software.  We delve into this issue further in Section~\ref{sec:fw} \luis{I dont see this in fw and I wouldr remove this sentence because it breaks the flow of the narrative}.

\luis{Amother thought: there are still many visualisations without asserts. We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors. this is already in the motivation but we can frame it as a visible open problem here as well.}

Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code. Very rarely assertions use dedicated API test methods. We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}

\luis{some thought that is blended here but maybe we want to make it clear. ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics). This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}

The wide adoption of ``magic'' thresholds (\luis{add numbers from results here}) is also concerning as it can hinder to the maintainability of ML projects. While in some cases there these thresholds come a results of domain expertise and experience acquired while working in a give project, we argue that such thresholds need to be properly documented. Our results show that there is a need to help practitioners keeping these decisions documented. This could be automated using a tool to flag undocumented decisions or that would suggest boilerplate documentation in Markdown or code annotations.

Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met. After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer. For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf. Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6). Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations. In other words, there is a need for automated tools that make this translation process easier for AI practitioners. In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.

@@

@@

In this study, we delved into the fusion of visualizations and analytical assertions in the testing of machine learning (ML) systems. Our investigation unveiled the emerging practice of crafting analytical assertions based on visual insights, a technique that bridges the qualitative and quantitative aspects of ML testing.

Testing ML systems in Jupyter Notebooks remains a burgeoning practice, with only a minority of notebooks adopting this approach. In total, we identified 269 pairs of visualizations and assertions that exhibited semantic correlation within 1.7k notebooks. Within these pairs, our analysis unveiled several key insights: nearly 40\% of the pairs included magic thresholds, 75\% of the assertions employed comparison operators, and 16\% of the assertions made use of external testing libraries. Furthermore, we observed that 22\% of the assertions did not utilize a comparison operator or a magic threshold.

Additionally, our study brought to light recurring testing patterns, underscoring the need yo enhance the alignment between visualizations and their corresponding assertions.
This research serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualizations and elevate the quality of analytical assertions.

\section{Future Work}\label{sec:fw}

This study underscores the imperative to enhance the current state of machine learning (ML) testing. Our observations have pointed to several potential future directions that could advance this field:

1. Tailored Assertion Tools: There is a clear need for assertion tools specifically designed to cater to the requirements of ML practitioners. These tools should provide assistance in identifying the most suitable visualizations for testing specific properties of ML systems. Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.

2. Automation of Assertions: The manual translation of insights derived from visualizations into analytical assertions presents a substantial challenge. Future work should focus on the development of automated tools that can assist AI practitioners in this translation process. By automating the generation of assertions from visual information, these tools can promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort.

3. Dynamic Testing in Real-World Scenarios: Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration. This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates. Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.

>>>>

  Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect $54,070$ Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part. Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems. And help ML practitioners formulate better analytical assertions.

@@

Visualisations are used throughout the ML development lifecycle to test various properties of the ML system. In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties. During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models. Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.

@@

We malysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation. Listing~\ref{lst:prompt} shows the prompt used for such cases. In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair. Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.

@@

% TODO add the 54K NB found with assert here as well

@@

\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq3}

@@

The results for RQ1 indicate that the combination of ML visualisations with analytical assertions is an emerging testing practice within the machine learning community. We found only $54,070$ notebooks that contain the keyword ``assert'', while there are more than a million Jupyter notebooks available on GitHub. This suggests that the practice of assertion is still relatively uncommon. We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties. We argue that guidelines on ML coding practices require improvement. And the need to steer education and research of Software Engineering practices towards the specific challenges faced  when contributing to ML software systems.

In the 1764 notebooks analysed in this study, several visualisations lack assertions. However, unformalised expectations still exist within each visualisation which cannot be readily validated by other participants. ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts. This highlights the need for refining the education for these professionals and crafting tools to compensate for potential gaps in their knowledge.

Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code. We hypothesise that this problem stems from two fundamental sources. The assertions in the VA pairs identified in this study lack dedicated testing libraries. This indicates that the current culture of testing ML code is still in its infancy and ML practitioners are not yet exploring software testing libraries. Second, the existing testing libraries are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.

% TODO \luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}

Results from RQ2 show the prevalent use of ``magic'' thresholds in assertions, which can impede the maintainability of ML projects. Even if these thresholds arise from domain expertise or experience obtained over time, it is crucial they are thoroughly documented. Our findings indicate a clear need to assist practitioners in documenting these decisions. Potential solutions include automated tools that highlight undocumented choices or suggest standardized documentation in Markdown or code annotations.

In the results from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. We argue that this can lead to a mismatch of expectations in future versions of the ML software. For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied. The examples presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial. Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of a ML developer.

For example, a linear correlation between the true and predicted labels is evident in Figure~{fig:r2}. Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric. Our results show that current software testing practices fail to capture the idiosyncrasies of ML code. There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.

@@

This study finds a clear need to improve the current ML testing process. There is a clear need for tools specifically designed to aid ML practitioners when testing ML systems. These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems. Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.

The manual translation of insights derived from visualisations into analytical assertions presents a substantial challenge. Future work should focus on the development of automated tools that can assist AI practitioners in this translation process. By automating the generation of assertions from visual information, these tools will promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort and bringing the code developed within the notebooks, closer to a production-ready codebase.

Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration. This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates. Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.

@@

This study study revealed the emerging practice of crafting analytical assertions based on visual insights to bridge the gap between qualitative and quantitative aspects of ML testing.

We identified 269 visualisation-assertion pairs from 1.7K Jupyter notebooks, that were semantically related to one another. And identified three recurring testing patterns within these pairs. Nearly 40\% of the assertions used ``magic'' thresholds, 75\% of the assertions used comparison operators, and 16\% of the assertions used external testing libraries.

This study serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualisations and elevate the quality of analytical assertions.
